---
title: "Calculating cutpoints"
output:
   md_document:
      variant: markdown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

[cutpointr](https://cran.r-project.org/web/packages/cutpointr/vignettes/cutpointr.html) can be used to calculate optimal cut offs.

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('cutpointr', 'randomForest')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
   }
  library(my_package, character.only = TRUE)
}

library(tidyverse)
theme_set(theme_bw())
```

## Example

The `suicide` data is a data frame with 532 rows and 4 variables:

1. `age` - (numeric) Age of participants in years
2. `gender` - (factor) Gender
3. `dsi` - (numeric) Sum-score (0 = low suicidality, 12 = high suicidality)
4. `suicide` - (factor) Past suicide attempt (no = no attempt, yes = at least one attempt)

```{r load_spam}
data(suicide)
head(suicide)
```

* `data` - A data.frame with the data needed for x, class and optionally subgroup.
* `x` - The variable name to be used for classification, e.g. predictions. The raw vector of values if the data argument is unused.
* `class` - The variable name indicating class membership. If the data argument is unused, the vector of raw numeric values.
* `method` - (function) A function for determining cutpoints. Can be user supplied or use some of the built in methods. See details.
* `metric` - (function) The function for computing a metric when using maximize_metric or minimize_metric as method and and for the out-of-bag values during bootstrapping. A way of internally validating the performance. User defined functions can be supplied, see details.

```{r cutpointr_suicide}
cp <- cutpointr(
  data = suicide,
  x = dsi,
  class = suicide, 
  method = maximize_metric,
  metric = sum_sens_spec
)

summary(cp)
```

Optimal cut off.

```{r optimal_cutoff}
cp$optimal_cutpoint
```

Plot.

```{r plot_cutpointr}
plot(cp)
```

[Youden's J statistic](https://en.wikipedia.org/wiki/Youden%27s_J_statistic) = sensitivity + specificity - 1.

```{r cutpointr_youden}
cp_youden <- cutpointr(
  data = suicide,
  x = dsi,
  class = suicide, 
  method = maximize_metric,
  metric = youden
)

summary(cp_youden)
```

## Random Forests

Use [spam data](https://archive.ics.uci.edu/ml/machine-learning-databases/spambase/spambase.names) to train a Random Forest model to test with `cutpointr`. Class 0 and 1 are ham (non-spam) and spam, respectively.

```{r random_forest_spam}
spam_data <- read.csv(file = "../data/spambase.csv")
spam_data$class <- factor(spam_data$class)
set.seed(1984)
rf <- randomForest(class ~ ., data = spam_data)
```

Optimal cut off using Youden's J statistic.

```{r spam_youden}
cp_rf_youden <- cutpointr(
  x = rf$votes[, 2],
  class = as.integer(spam_data$class)-1,
  method = maximize_metric,
  metric = youden
)

summary(cp_rf_youden)
```

Plot.

```{r plot_spam_youden}
plot(cp_rf_youden)
```

Plot metric.

```{r plot_metric}
plot_metric(cp_rf_youden)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
