---
title: "Evaluation formulae"
output:
   md_document:
      variant: markdown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

Mathematical formulae elaborated.

## Root Mean Square Error

The Root Mean Square Error (RMSE) is often used in regression problems to indicate how much error was made by the predictions, with a higher value to larger errors.

$$ RMSE(X, h) = \sqrt{ \frac{1}{m} \sum^m_{i = 1} ( h(x^{(i)}) - y^{(i)} )^2 } $$

* $m$ is the number of instances/cases in the dataset
* $x^{(i)}$ is a vector of all the feature values of the $i^{th}$ instance in the dataset and $y^{(i)}$ is the label
* $X$ is a matrix containing all the feature values of all instances in the dataset. There is one row per instance, and the $i^{th}$ row is equal to the transpose of $x^{(i)}$, noted $(x^{(i)})^T$.
* $h$ is the prediction function, also called a hypothesis. When the function is given an instance's feature vector $x^{(i)}$, it outputs a predicted value $\hat{y}^{(i)} = h(x^{(i)})$ for that instance.
* $RMSE(X,h)$ is the cost function measured on the set using hypothesis $h$.

## Mean Absolute Error

If there are many outliers, the Mean Absolute Error (MAE, also called the average absolute deviation) can be considered.

$$ MAE(X, h) = \frac{1}{m} \sum^m_{i = 1} |h(x^{(i)}) - y^{(i)}| $$

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
