---
title: "k-Nearest Neighbours"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

In pattern recognition, the [k-Nearest Neighbours algorithm](https://en.wikipedia.org/wiki/K-nearest_neighbors_algorithm) (k-NN) is a non-parametric method used for classification and regression. In both cases, the input consists of the k closest training examples in the feature space.

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- 'FNN'

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}
```

## Regression

We'll use the `women` dataset to demonstrate how k-NN performs regression. The dataset contains height and weight measurements for 15 American women aged between 30–39.

```{r}
data(women)
str(women)
plot(women, xlab = "Height (in)", ylab = "Weight (lb)", main = "women data: American women aged 30-39", pch = 16)
```

In the example below, we want to *predict the weight* of a female who is 60 inches tall based on data in the `women` dataset.

```{r}
knn <- function(x, x_train, y_train, k){
  d <- abs(x - x_train)
  s <- order(d)
  return(mean(y_train[s[1:k]]))
}

# using four neighbours
knn(60, women$height, women$weight, 4)

# using five neighbours
knn(60, women$height, women$weight, 5)
```

* The `knn` algorithm first calculates the absolute distance of an input to a known set of data points for the same variable (height).
* These distances are then sorted, with the closest data points ranked first.
* The k-nearest distances of heights are used to obtain the corresponding weights
* Finally, the k weights are averaged (mean) and returned

Use `sapply` to predict several values.

```{r}
sapply(c(60,70), knn, x_train = women$height, y_train = women$weight, k = 4)
```

You can also use `knn.reg` in the `FNN` package.

```{r}
knn.reg(women$height, 60, women$weight, 4)

knn.reg(women$height, 60, women$weight, 5)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

