---
title: "MNIST"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('keras', 'tensorflow')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}

# install_tensorflow()
```

## MNIST dataset

```{r}
mnist <- dataset_mnist()
train_images <- mnist$train$x
train_labels <- mnist$train$y
test_images <- mnist$test$x
test_labels <- mnist$test$y
```

```{r}
network <- keras_model_sequential() %>%
   layer_dense(units = 512, activation = "relu", input_shape = c(28 * 28)) %>%
   layer_dense(units = 10, activation = "softmax")
```

```{r}
network %>% compile(
   optimizer = "rmsprop",
   loss = "categorical_crossentropy",
   metrics = c("accuracy")
)
```

```{r}
train_images <- array_reshape(train_images, c(60000, 28 * 28))
train_images <- train_images / 255
test_images <- array_reshape(test_images, c(10000, 28 * 28))
test_images <- test_images / 255
train_labels <- to_categorical(train_labels)
test_labels <- to_categorical(test_labels)
```

```{r}
network %>% fit(train_images, train_labels, epochs = 5, batch_size = 128)
```

```{r}
metrics <- network %>% evaluate(test_images, test_labels)
metrics
```

```{r}
network %>% predict(test_images[1:10,]) %>% k_argmax()
```

## Fashion MNIST

Following the [Basic Image Classification](https://tensorflow.rstudio.com/tutorials/beginners/basic-ml/tutorial_basic_classification/) guide.

This guide uses the Fashion MNIST dataset which contains 70,000 grayscale images in 10 categories. The images show individual articles of clothing at low resolution (28 by 28 pixels).

Fashion MNIST is intended as a drop-in replacement for the classic MNIST dataset and is a slightly more challenging problem than regular MNIST.

We will use 60,000 images to train the network and 10,000 images to evaluate how accurately the network learned to classify images.

```{r fashion_mnist}
fashion_mnist <- dataset_fashion_mnist()

str(fashion_mnist)
```

Assign.

```{r fashion_mnist_data}
c(train_images, train_labels) %<-% fashion_mnist$train
c(test_images, test_labels) %<-% fashion_mnist$test
```

The labels are arrays of integers ranging from 0 to 9 and correspond to the class of clothing the image represents:

| Digit |    Class    |
| ----- | ----------- |
|   0   | T-shirt/top |
|   1   | Trouser     |
|   2   | Pullover    |
|   3   | Dress       |
|   4   | Coat        |
|   5   | Sandal      |
|   6   | Shirt       |
|   7   | Sneaker     |
|   8   | Bag         |
|   9   | Ankle boot  |

```{r class_names}
class_names = c(
   'T-shirt/top',
   'Trouser',
   'Pullover',
   'Dress',
   'Coat', 
   'Sandal',
   'Shirt',
   'Sneaker',
   'Bag',
   'Ankle boot'
)
```

```{r}
image_1 <- as.data.frame(train_images[1, , ])
colnames(image_1) <- seq_len(ncol(image_1))
image_1$y <- seq_len(nrow(image_1))
image_1 <- gather(image_1, "x", "value", -y)
image_1$x <- as.integer(image_1$x)

ggplot(image_1, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "black", na.value = NA) +
  scale_y_reverse() +
  theme_minimal() +
  theme(panel.grid = element_blank())   +
  theme(aspect.ratio = 1) +
  xlab("") +
  ylab("")
```

Scale values to a range of 0 to 1 by dividing by 255 before feeding to the neural network model. It is important that the training set and the testing set are pre-processed in the same way.

```{r}
train_images <- train_images / 255
test_images <- test_images / 255
```

Display the first 25 images.

```{r}
par(mfcol=c(5,5))
par(mar=c(0, 0, 1.5, 0), xaxs='i', yaxs='i')
for (i in 1:25) { 
  img <- train_images[i, , ]
  img <- t(apply(img, 2, rev)) 
  image(1:28, 1:28, img, col = gray((0:255)/255), xaxt = 'n', yaxt = 'n',
        main = paste(class_names[train_labels[i] + 1]))
}
```

Set up the layers.

```{r}
model <- keras_model_sequential()
model %>%
  layer_flatten(input_shape = c(28, 28)) %>%
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dense(units = 10, activation = 'softmax')
```

```{r}
model %>% compile(
  optimizer = 'adam', 
  loss = 'sparse_categorical_crossentropy',
  metrics = c('accuracy')
)
```

```{r}
model %>% fit(train_images, train_labels, epochs = 5, verbose = 2)
```

```{r}
score <- model %>% evaluate(test_images, test_labels, verbose = 0)
score
```

```{r}
predictions <- model %>% predict(test_images)
predictions[1, ]
```
```{r}
which.max(predictions[1, ])
```

```{r}
class_pred <- model %>% predict(test_images) %>% k_argmax()

class_pred[1:20]
```

```{r}
test_labels[1:20]
```

```{r}
par(mfcol=c(5,5))
par(mar=c(0, 0, 1.5, 0), xaxs='i', yaxs='i')
for (i in 1:25) { 
   img <- test_images[i, , ]
   img <- t(apply(img, 2, rev)) 
   # subtract 1 as labels go from 0 to 9
   predicted_label <- which.max(predictions[i, ]) - 1
   true_label <- test_labels[i]
   if (predicted_label == true_label) {
      color <- '#008800' 
   } else {
      color <- '#bb0000'
   }
   image(1:28, 1:28, img, col = gray((0:255)/255), xaxt = 'n', yaxt = 'n',
         main = paste0(class_names[predicted_label + 1], " (",
                       class_names[true_label + 1], ")"),
         col.main = color)
}
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
