---
title: "Logistic regression"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

Logistic regression is a regression model where the dependent variable (DV) is categorical. Example from Wikipedia: "Logistic regression may be used to predict whether a patient has a given disease (e.g. diabetes; coronary heart disease), based on observed characteristics of the patient (age, sex, body mass index, results of various blood tests, etc.)."

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('Amelia', 'ROCR')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}
```

## Hours of study

Using the example from Wikipedia: [Probability of passing an exam versus hours of study](https://en.wikipedia.org/wiki/Logistic_regression#Example:_Probability_of_passing_an_exam_versus_hours_of_study)

```{r}
d <- data.frame(
   hours = c(0.50,0.75,1.00,1.25,1.50,1.75,1.75,2.00,2.25,2.50,2.75,3.00,3.25,3.50,4.00,4.25,4.50,4.75,5.00,5.50),
   pass = factor(c(0,0,0,0,0,0,1,0,1,0,1,0,1,0,1,1,1,1,1,1))
)

model <- glm(
   pass ~ hours,
   family=binomial(link='logit'),
   data=d
)
summary(model)
```

The output indicates that hours studying is significantly associated with the probability of passing the exam (p=0.0167, Wald test). The output also provides the coefficients for Intercept = -4.0777 and Hours = 1.5046.

Probability of passing as a function of hours of study.

```{r}
prob_passing <- function(hours){
  1 / (1 + exp(-(-4.0777 + 1.5046 * hours)))
}

prob_passing(4)
```

## Survival on the Titanic

Adapted from [How to Perform a Logistic Regression in R](https://datascienceplus.com/perform-logistic-regression-in-r/).

```{r}
data <- read.csv("../data/titanic.csv.gz", na.strings = '')

# missing data
sapply(data, function(x) sum(is.na(x)))

missmap(data)
```

Remove some features

```{r}
data_subset <- select(data, -PassengerId, -Ticket, -Cabin, -Name)

# remove the two cases with missing embarked data
data_subset <- filter(data_subset, !is.na(Embarked))

# you can use the mean age for the missing ages
data_subset$Age[is.na(data_subset$Age)] <- mean(data_subset$Age, na.rm=TRUE)

# subset into training and testing sets
train <- data_subset[1:800,]
test  <- data_subset[801:nrow(data_subset),]

model <- glm(Survived ~.,
             family=binomial(link='logit'),
             data=train)

fitted <- predict(model,
                  newdata=test[,-1],
                  type='response')
```

```{r}
pr <- prediction(fitted, test$Survived, )
prf <- performance(pr, measure = "tpr", x.measure = "fpr")
auc <- performance(pr, measure = "auc")
plot(prf)
legend(x = 0.75, y = 0.05, legend = paste("AUC = ", auc@y.values), bty = 'n')
```

## Links

* [Simple logistic regression](http://www.biostathandbook.com/simplelogistic.html)
* [Multiple logistic regression](http://www.biostathandbook.com/multiplelogistic.html)

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

