---
title: "Support Vector Machine"
output:
  pdf_document: default
  html_notebook: default
---

Using the [Breast Cancer Wisconsin (Diagnostic) Data Set](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)).

```{r}
my_link <- 'http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data'
data <- read.table(url(my_link), stringsAsFactors = FALSE, header = FALSE, sep = ',')
names(data) <- c('id','ct','ucsize','ucshape','ma','secs','bn','bc','nn','miti','class')
head(data)
```

Any missing data?

```{r}
dim(data)[1] * dim(data)[2]
table(is.na.data.frame(data))
```

What's the structure?

```{r}
str(data)
```

Why is the bare nuclei (bn) stored as characters instead of integers?

```{r}
table(data$bn)
```

Change the question marks into NA's and then into median values.

```{r}
data$bn <- gsub(pattern = '\\?', replacement = NA, x = data$bn)
data$bn <- as.integer(data$bn)
my_median <- median(data$bn, na.rm = TRUE)
data$bn[is.na(data$bn)] <- my_median
str(data)
```

The class should be a factor; 2 is benign and 4 is malignant.

```{r}
data$class <- factor(data$class)
```

Finally remove id the row name, which was not unique anyway.

```{r}
data <- data[,-1]
```

Separate into training (80%) and testing (20%).

```{r}
set.seed(31)
my_decider <- rbinom(n=nrow(data),size=1,p=0.8)
table(my_decider)
train <- data[as.logical(my_decider),]
test <- data[!as.logical(my_decider),]
```

Using the `e1071` package.

```{r}
library(e1071)

tuned <- tune.svm(class ~ ., data = train, gamma = 10^(-6:-1), cost = 10^(-1:1))
summary(tuned)
```

Train model using the best values for gamma and cost.

```{r}
svm_model <- svm(class ~ ., data = train, kernel="radial", gamma=0.01, cost=1)
summary(svm_model)
```

Predict test cases.

```{r}
svm_predict <- predict(svm_model, test)
table(svm_predict, test$class)
```


