---
title: "Breast cancer data"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

Using the [Breast Cancer Wisconsin (Diagnostic) Data Set](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)).

```{r save_data}
my_link <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data'
data <- read.table(url(my_link), stringsAsFactors = FALSE, header = FALSE, sep = ',')
names(data) <- c('id','ct','ucsize','ucshape','ma','secs','bn','bc','nn','miti','class')
head(data)
```

Any missing data?

```{r any_missing_data}
any(is.na(data))
```

Data structure.

```{r str}
str(data)
```

Bare nuclei (`bn`) was stored as characters because of question marks.

```{r bn}
table(data$bn)
```

Change the question marks into NA's and then into median values.

```{r convert_na}
data$bn <- as.integer(gsub(pattern = '\\?', replacement = NA, x = data$bn))

data$bn[is.na(data$bn)] <- median(data$bn, na.rm = TRUE)

str(data)
```

Convert `class` into a factor with levels `2` for benign and `4` for malignant.

```{r factor_class}
data$class <- factor(data$class)
```

The `id` column is duplicated and it is not clear why they are duplicated.

```{r dup_data}
dup_idx <- which(duplicated(data$id))
dup_data <- data[data$id %in% data[dup_idx, 'id'], ]

head(dup_data[order(dup_data$id), ])
```

We will remove duplicated IDs.

```{r remove_dup_id}
dim(data)

data <- data[! data$id %in% data[dup_idx, 'id'], ]

dim(data)

write.csv(
  x = data,
  file = "breast_cancer_data.csv",
  quote = FALSE,
  row.names = FALSE
)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

