---
title: "Spambase Data Set"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

From <https://archive.ics.uci.edu/ml/datasets/spambase>

>Our collection of spam e-mails came from our postmaster and individuals who had filed spam. Our collection of non-spam e-mails came from filed work and personal e-mails, and hence the word 'george' and the area code '650' are indicators of non-spam. These are useful when constructing a personalized spam filter. One would either have to blind such non-spam indicators or get a very wide collection of non-spam to generate a general purpose spam filter.

The last column of `spambase.data` denotes whether the e-mail was considered spam (1) or not (0). Most of the attributes indicate whether a particular word or character was frequently occurring in the e-mail.  The run-length attributes (55-57) measure the length of sequences of consecutive capital letters.

```{r save_data}
data_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/spambase/spambase.data'
data <- read.table(url(data_url), stringsAsFactors = FALSE, header = FALSE, sep = ',')

data_col_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/spambase/spambase.names'
data_col <- read.table(url(data_col_url), stringsAsFactors = FALSE, header = FALSE, comment.char = "|")

my_cols <- c(gsub(":$", "", data_col$V1[-1]), 'class')

colnames(data) <- my_cols
dim(data)
```

Any missing data?

```{r any_missing_data}
any(is.na(data))
```

Data structure.

```{r str}
str(data)
```

Convert `class` into a factor with levels `0` for ham and `1` for spam.

```{r factor_class}
data$class <- factor(data$class)
```

Save as CSV.

```{r save_csv}
write.csv(
  x = data,
  file = "spambase.csv",
  quote = FALSE,
  row.names = FALSE
)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
