---
title: "Hierarchical clustering"
output:
   md_document:
      variant: markdown
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

Here's a [good post](http://datascienceplus.com/hierarchical-clustering-in-r/) on hierarchical clustering. Using data from the `datamicroarray` package.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('dendextend', 'remotes', 'datamicroarray')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      if (my_package == 'datamicroarray'){
         install_github('ramhiser/datamicroarray')
      } else {
         install.packages(my_package, '/packages')
      }
      library(my_package, character.only = TRUE)
   }
}
```

## Using microarray data

I will use the `yeoh` data set.

```{r}
data('yeoh', package = "datamicroarray")

dim(yeoh$x)

table(yeoh$y)
```

Calculate distance between all samples.

```{r}
choose(248, 2)

my_dist <- dist(yeoh$x)

summary(my_dist)
```

Perform hierarchical clustering using complete (maximum) linkage, which is the default.

```{r}
my_hclust <- hclust(my_dist)
```

Form six clusters based on the clustering.

```{r}
my_clus <- cutree(my_hclust, k = 6)

table(my_clus, yeoh$y)

cluster_one <- yeoh$y[my_clus == 1]
```

Form `n` clusters based on arbitrary distance.

```{r}
my_clus_two <- cutree(my_hclust, h = 25)
# much more homogeneous
table(my_clus_two, yeoh$y)
```

Plot.

```{r}
my_hclust_mod <- my_hclust
my_hclust_mod$labels <- as.vector(yeoh$y)
plot(color_branches(my_hclust_mod, h = 25, groupLabels = TRUE))
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

