---
title: "Artificial Neural Networks"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('MASS', 'neuralnet')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}
```

## Compare neural network regression with linear regression

Following the example from [this tutorial](http://datascienceplus.com/fitting-neural-network-in-r/).

```{r}
set.seed(500)
data <- Boston
apply(data,2,function(x) sum(is.na(x)))

index <- sample(1:nrow(data), round(0.75*nrow(data)))
train <- data[index,]
test <- data[-index,]
lm.fit <- glm(medv ~ ., data=train)
summary(lm.fit)
pr.lm <- predict(lm.fit, test)
MSE.lm <- sum((pr.lm - test$medv)^2)/nrow(test)
```

## Feature scaling

[Feature scaling](https://en.wikipedia.org/wiki/Feature_scaling) using:

$$ x' = \frac{x - min(x)}{max(x) - min(x)} $$

```{r}
x <- 1:20
x_a <- (x - min(x)) / (max(x) - min(x))
x_b <- as.vector(scale(x, center = min(x), scale = max(x) - min(x)))
x_a
identical(x_a, x_b)
```

Carrying out the normalisation.

```{r}
maxs <- apply(data, 2, max)
mins <- apply(data, 2, min)

scaled <- as.data.frame(scale(data, center = mins, scale = maxs - mins))

train_ <- scaled[index,]
test_ <- scaled[-index,]
```

## Training

Manually create formula as `f` since neuralnet() doesn't recognise `medv ~ .`.

```{r}
n <- names(train_)
f <- as.formula(paste("medv ~", paste(n[!n %in% "medv"], collapse = " + ")))

nn <- neuralnet(f, data = train_, hidden=c(5,3), linear.output = TRUE)
plot(nn)
```

## Prediction

We need to unscale the predictions and test; remember the formula:

$$ x' = \frac{x - min(x)}{max(x) - min(x)} $$

```{r}
x <- 1:20
x_a <- (x - min(x)) / (max(x) - min(x))
x_orig <- as.integer(x_a * (max(x) - min(x)) + (min(x)))
identical(x, x_orig)
```

Compare RMS of neural network regression with linear regression.

```{r}
pr.nn <- compute(nn, test_[,1:13])
pr.nn_ <- pr.nn$net.result * (max(data$medv) - min(data$medv)) + min(data$medv)
test.r <- (test_$medv) * (max(data$medv) - min(data$medv)) + min(data$medv)

MSE.nn <- sum((test.r - pr.nn_)^2)/nrow(test_)
print(paste(MSE.lm,MSE.nn))
```

## Further reading

The neuralnet [reference manual](https://cran.r-project.org/web/packages/neuralnet/neuralnet.pdf).

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

