---
title: "Decision trees"
output:
   md_document:
      variant: markdown
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

A [decision tree](https://en.wikipedia.org/wiki/Decision_tree) is a decision support tool that uses a tree-like graph or model of decisions and their possible consequences, including chance event outcomes, resource costs, and utility.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('tree', 'rpart', 'rpart.plot')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}
```


```{r}
tree1 <- tree(Species ~ Sepal.Width + Petal.Width, data = iris)
summary(tree1)
plot(tree1)
text(tree1)
```

```{r}
plot(iris$Petal.Width,
     iris$Sepal.Width,
     pch=19,
     col=as.numeric(iris$Species))

partition.tree(tree1, label="Species", add=TRUE)
legend(2.3,4.5,
       legend=levels(iris$Species),
       col=1:length(levels(iris$Species)),
       pch=19,
       bty = 'n')
```

```{r}
tree2 <- tree(Species ~ ., data = iris)
summary(tree2)
plot(tree2); text(tree2)
```

Each node shows:

1. The predicted class (setosa, versicolor, and virginica)
2. The numbers of each class (in the order above)
3. The percentage of all samples

```{r}
rpart <- rpart(Species ~ ., data=iris, method="class")
summary(rpart)
rpart.plot(rpart, type = 4, extra = 101)
```

## Titanic data

```{r}
titanic <- read.csv('../data/titanic.csv.gz')
str(titanic)

titanic$Pclass <- factor(titanic$Pclass)
boxplot(Fare ~ Pclass, data = titanic)
```

Each node shows:

1. The predicted class (0 or 1)
2. The predicted probability of survival
3. The percentage of all samples

```{r}
t <-  rpart(Survived ~ Sex + Fare + Age, data=titanic, method="class")
rpart.plot(t)
```

For example the 0.74 indicates that 74% of females survived.

```{r}
prop.table(
   table(
      titanic$Sex, titanic$Survived
   ), margin = 1
)
```

## Breast cancer data

```{r}
data <- read.table(
   "../data/breast_cancer_data.csv",
   stringsAsFactors = FALSE,
   sep = ',',
   header = TRUE
)
data$class <- factor(data$class)
data <- data[,-1]
```

Each node shows:

1. The predicted class (0 or 1)
2. The predicted probability of malignancy
3. The percentage of all samples

```{r}
t <- rpart(class ~ ., data = data, method="class")
rpart.plot(t)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

