---
title: "Proximus"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_bw())
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
```

## Introduction

https://en.wikibooks.org/wiki/Data_Mining_Algorithms_In_R/Clustering/Proximus

Install packages if missing and load.

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- 'cba'

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
      library(my_package, character.only = TRUE)
   }
}
```

## Results

https://www.rdocumentation.org/packages/cba/versions/0.2-21/topics/proximus

x: a logical matrix. 
max.radius: the maximum number of bits a member in a row set may deviate from its dominant pattern. 
min.size: the minimum split size of a row set. 
min.retry: number of retries to split a pure rank-one approximation (translates into a resampling rate). 
max.iter: the maximum number of iterations for finding a local rank-one approximation. 
debug: optional debugging output.

```{r}
x <- rlbmat()
pr <- proximus(x, max.radius=8, debug=TRUE)
op <- par(mfrow=c(1,2), pty="s")
lmplot(x, main="Data")
box()
lmplot(fitted(pr)$x, main="Approximation")
box()
par(op)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```

