---
title: "Gaussian Mixture Models"
output:
   md_document:
      variant: markdown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
options(scipen = 999)
library(tidyverse)
theme_set(theme_bw())
```

## Introduction

Following these blog posts:

* <https://tinyheero.github.io/2015/10/13/mixture-model.html>
* <https://tinyheero.github.io/2016/01/03/gmm-em.html>
* <https://exploringdatablog.blogspot.com/2011/08/fitting-mixture-distributions-with-r.html>

Install and load [mixtools](https://cran.r-project.org/web/packages/mixtools/index.html).

```{r load_package, message=FALSE, warning=FALSE}
.libPaths('/packages')
my_packages <- c('mixtools')

for (my_package in my_packages){
   if(!require(my_package, character.only = TRUE)){
      install.packages(my_package, '/packages')
   }
  library(my_package, character.only = TRUE)
}
```

Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.

A data frame with 272 observations on 2 variables.

* `eruptions` - Eruption time in minutes
* `waiting` - Waiting time to next eruption in minutes

```{r faithful_scatter}
ggplot(faithful, aes(waiting, eruptions)) +
  geom_point()
```

The `normalmixEM` function builds a 2-component GMM (k = 2 indicates to use 2 components).

```{r mixmdl}
set.seed(1984)
mixmdl <- normalmixEM(faithful$waiting, k = 2)
summary(mixmdl)
```

The red and blue lines indicate the two fitted Gaussian distributions.

```{r plot_density}
#' Plot a Mixture Component
#' 
#' @param x Input data
#' @param mu Mean of component
#' @param sigma Standard deviation of component
#' @param lam Mixture weight of component
plot_mix_comps <- function(x, mu, sigma, lam) {
  lam * dnorm(x, mu, sigma)
}

data.frame(x = mixmdl$x) %>%
  ggplot() +
  geom_histogram(aes(x, ..density..), binwidth = 1, colour = "black", 
                 fill = "white") +
  stat_function(geom = "line", fun = plot_mix_comps,
                args = list(mixmdl$mu[1], mixmdl$sigma[1], lam = mixmdl$lambda[1]),
                colour = "red", lwd = 1.5) +
  stat_function(geom = "line", fun = plot_mix_comps,
                args = list(mixmdl$mu[2], mixmdl$sigma[2], lam = mixmdl$lambda[2]),
                colour = "blue", lwd = 1.5) +
  ylab("Density")
```

Mean and standard deviation.

```{r parameters}
mixmdl$mu
mixmdl$sigma
```

`lambda` indicates the ratio of the dataset but formally it is referred to as the mixing weights (or mixing proportions or mixing co-efficients). It can be interpreted as the red component representing 36% and the blue component representing 64% of the input data.

```{r lambda}
mixmdl$lambda
sum(mixmdl$lambda)
```

Each input data point is assigned a posterior probability of belonging to one of these components.

```{r pp}
head(cbind(mixmdl$x, mixmdl$posterior))
```

## Predict

When two distributions are close to each other.

```{r close}
set.seed(1984)
k1 <- rnorm(n = 123, mean = 0.4, sd = 0.2)
k2 <- rnorm(n = 156, mean = 0.6, sd = 0.2)

plot(density(c(k1, k2)), main = '')
```

EM.

```{r normalmixem_close}
set.seed(1984)
mm <- normalmixEM(c(k1, k2), k = 2, maxit = 10000)
summary(mm)
```

Two more distinguishable distributions.

```{r distinct}
set.seed(1984)
k1 <- rnorm(n = 123, mean = 0.3, sd = 0.2)
k2 <- rnorm(n = 156, mean = 0.7, sd = 0.2)

df <- tibble(
  label = rep(c('k1', 'k2'), c(length(k1), length(k2))),
  value = c(k1, k2)
)

ggplot(df, aes(value, fill = label)) +
  geom_histogram(bins = 25, position = "dodge")
```

EM.

```{r normalmixem_distinct}
set.seed(1984)
mm2 <- normalmixEM(df$value, k = 2, maxit = 10000)
summary(mm2)
```

Component 1 is the distribution with the higher mean (`k2`).

```{r confusion_matrix}
table(
  real = df$label,
  predicted = ifelse(mm2$posterior[, 1] < 0.5, yes = 'k1', no = 'k2')
)
```

Two clearly distinguishable distributions.

```{r clearly_distinct}
set.seed(1984)
k1 <- rnorm(n = 123, mean = 0.2, sd = 0.2)
k2 <- rnorm(n = 156, mean = 0.8, sd = 0.2)

df2 <- tibble(
  label = rep(c('k1', 'k2'), c(length(k1), length(k2))),
  value = c(k1, k2)
)

ggplot(df2, aes(value, fill = label)) +
  geom_histogram(bins = 25, position = "dodge")
```

EM and predict.

```{r normalmixem_clearly_distinct}
set.seed(1984)
mm3 <- normalmixEM(df2$value, k = 2, maxit = 10000)
table(
  real = df2$label,
  predicted = ifelse(mm3$posterior[, 2] < 0.5, yes = 'k1', no = 'k2')
)
```

Do we do better by simply setting a threshold at 0.5? In this case, yes.

```{r hard_filter}
table(
  real = df2$label,
  predicted = ifelse(df2$value < 0.5, yes = 'k1', no = 'k2')
)
```

## Session info

Time built.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
